A Method for the Spectroscopic Inference of Stellar Parameters

### 2014-07-06 tex/paper.tex
PHOENIX regions

We adopt a somewhat analogous approach to deal with this issue, but cast it internal to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  This is achieved in practice by distorting the 
model spectrum with a (low-order) Chebyshev polynomial \citep[e.g.,][]{eisenstein06,koleva09}, 
\begin{equation}
\vM(\vt, \vT_{\rm cheb}) = \vM(\vt) \times \sum_n c_n \, T_n,
\end{equation}
where $T_n$ are the standard Chebyshev functions of order $n$ and the coefficients are treated as a
set of nuisance (hyper-)parameters, $\vT_{\rm cheb} = \{c_n\}^{\rm order}$, for each available 
spectral order in the dataset.  With judicious priors on $\vT_{\rm cheb}$, we can ensure that the 
unintended treatment of real spectral features (e.g., broad, deep molecular bands) as calibration 
artifacts is negligible (see Section~\ref{sec:examples} for examples).  The lowest-order (scaling) 
coefficient, $c_0$, is by its nature degenerate with the solid angle parameter, $\Omega$.  
Therefore, we enforce an additional constraint by requiring that the mean of the polynomial is 
unity.  For data with a single spectral order, this means setting $c_0 = 1$.  If the goal is to 
model multiple spectral orders, we assign $c_0 = 1$ in an arbitrary order as an anchor, but permit 
the $c_0$ values in other orders to be different (as necessary).  It is worth noting that this 
formalism can, in principle, be extended to develop models for completely uncalibrated spectra, 
rather than the residual calibration mismatches as described here (with a suitable relaxation of 
the priors on $\vT_{\rm cheb}$).  Figure \ref{fig:chebyshev} offers a practical demonstration of how 
these nuisance parameters are applied.

\begin{figure}[!htb]
\begin{center}
  \includegraphics{figs/chebyshev.pdf}
  \caption{A region of the echelle spectrum at the overlap between two adjacent orders of the echelle spectrum. Panel 1: In green and blue, we have the flux-calibrated dataset. The adjacent orders overlap but there is a slight discrepancy of $\lesssim 3$\%. In the background in red is a model spectrum generated for reasonable values of stellar parameters, $\vt$. Panel 2: In order to account for residual errors in the flux calibration, we allow the model to be multiplied by a Chebyshev polynomial. Here are the polynomials determined as best fit, to multiply the model. Panels 3 \& 4: In principle, we could divide the data by these polynomials to recover what the true flux-calibrated data should be (panel 3), but instead we multiply the model by the polynomials to recover the original discrepancy between orders (panel 4). This has the advantage of preserving the original dataset as a fixed quantity and makes the model linear in the Chebyshev polynomial coefficients.}
\label{fig:chebyshev}
\end{center}
\end{figure}

@@

Assuming that these local residuals are produced primarily by pathological differences in the 
spectral line strength (rather than shape or center), a simple Gaussian is a reasonable residual 
model.  In that case, the $k^{\rm th}$ such local residual can be described as
\begin{equation}
\vR_{\lambda}(a_k, \mu_k, \sigma_k) = \frac{a_k}{\sqrt{2\pi}{\sigma_k}} \exp \left[ - \frac{r^2(\lambda,\mu_k)}{2\sigma_k^2} \right],
\end{equation}
with an amplitude $a_k$, a mean wavelength $\mu_k$, and a width $\sigma_k$.  Following 
Eq.~\ref{eqn:expectation}, the kernel that describes the covariance of any two pixels related to 
the $k^{\rm th}$ residual region is
\begin{equation}
  \mathcal{K}_k(\lambda_i, \lambda_j | a_k, \mu_k, \sigma_k) = \frac{1}{2\pi} \left(\frac{a_k}{\sigma_k}\right)^2 \exp \left [ - \, \frac{r^2(\lambda_i, \mu_k) + r^2(\lambda_j, \mu_k)}{2 \sigma_k^2}\right ].
\label{eqn:klocal}
\end{equation}
We again taper this kernel with a Hann window (Eq.~\ref{eqn:Hann}) to ensure computational 
efficiency with a sparse covariance matrix; in this case, the truncation distance $r_0$ can be set 
to some multiple of the Gaussian width (we set $r_0 = 4\sigma_k$).  
Figure \ref{fig:region} demonstrates how this non-stationary Gaussian 
kernel generates a localized region of enhanced variance that successfully mimics the kind of 
residuals produced by an inaccurate spectral line model.  In effect, these kernels down-weight the 
influence of such strong residuals in the likelihood calculation, mitigating any potential bias 
they might induce on inferences of the interesting parameters ($\vt$).  In essence, this is a 
robust, flexible, and unbiased method for (correlated) outlier rejection that preserves the 
integrity of the probabilistic framework being developed (as opposed to manual or threshold-based 
clipping/masking).

@@

We can now compute the covariance matrix employed in the likelihood calculation 
(Eq.~\ref{eqn:lnprob}) as the linear combination of these kernels and the trivial pixel-by-pixel 
noise matrix, 
\begin{eqnarray}
\vC_{ij}(\vT_{\rm cov}) & = & b \, \delta_{ij}\sigma_i + \\ \nonumber
         &   &  w(r_{ij}|r_0 = 4\ell) \,\, \mathcal{K}_g(\lambda_i, \lambda_j|a_g, \ell) + \\ \nonumber
         &   &  \sum_k w(r_{ij}|r_0 = 4\sigma_k) \,\, \mathcal{S}_k(\lambda_i,\lambda_j | h_k) \,\, \mathcal{K}_k(\lambda_i, \lambda_j | a_k, \mu_k, \sigma_k),
\end{eqnarray}
where the covariance hyperparameters $\vT_{\rm cov} = [a_g, \ell, \{a_k, \mu_k, \sigma_k, 
h_k\}^{N_{\rm reg}}]$, $N_{\rm reg}$ is the number of local residual regions
(see below for details on how this is determined), and $b$ is assumed to be a
fixed parameter that scales up the pixel noise values to account for read
noise, noise added during the 2D to 1D spectral extraction procedure, and
interpolation uncertainties (see Section \ref{subsec:synthetic}; reasonable
values are $b \approx 1.02$--1.10 for well-calibrated optical spectra; see Section
\ref{sec:examples} for examples).

@@

\comm{Not much done here...want to meet and discuss before I try editing.}

@@

\todo{We can (should?) add an additional level to the hierarchy of
  hyperparameters and add parameters to describe the \emph{population
  characteristics} of poorly modeled spectral lines (mostly the typical width,
  amplitude of these lines).  This will tell us about the frequency and
distribution of spectral modelling errors.} We explicitly fit for the
hyperparameters of the covariance kernels at the same time we fit for the
stellar parameters. While including these extra parameters does increase the
dimensionality and complexity of our model, they do provide several advantages.
\emph{A priori} we do not know which regions of the spectrum are improperly
modeled. The covariance hyperparameters provide a framework to identify these
regions iteratively and in a self-consistent manner.

@@

@@

\begin{deluxetable}{llrlr@{$\pm$}lr@{$\pm$}lr@{$\pm$}ll} 
\tablecaption{\label{table:tests} Tests}
\tablehead{
  \colhead{Object} & \colhead{Orders} & \colhead{$\lambda$ range [\AA]} & \colhead{library} &
  \multicolumn{2}{c}{$T_{\rm eff} \pm \sigma$} & \multicolumn{2}{c}{$\log g \pm \sigma$} &
  \multicolumn{2}{c}{$\Z \pm \sigma$} & \colhead{comments}
}
\startdata
WASP-14 & 22-24 & 5060-5315 & Kurucz & 6310 & 52 & 3.91  & 0.08 & -0.38  & 0.03 & Poisson only\\
WASP-14 & 22-24 & 5060-5315& Kurucz & 6314 & 50 & 3.92 & 0.08 & -0.39 & 0.03 & Matern only\\
WASP-14 & 22-24 & 5060-5315& Kurucz & 6512 & 32 & \multicolumn{2}{c}{\nodata} & -0.27 & 0.02 & Matern + fixed $\log g = 4.29$\\
WASP-14 & 23 & & PHOENIX & 6021 & 16 & 3.83 & 0.3 & -0.50 & 0.01 & Poisson only \\
WASP-14 & 23 & & PHOENIX & & & & & & & Matern only, no regions \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & 6117 & 30 & 3.73 & 0.06 & -0.52 & 0.02 & Poisson only \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & 5865 & 43 & 3.2 & 0.08 & -0.85 & 0.03 & Matern only. Because of 24. \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & & & \multicolumn{2}{c}{\nodata} & & & Matern + fixed $\log g = 4.29$ \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & & & & & & & Regions included \\
\enddata
\end{deluxetable}

>>>>

We adopt a somewhat analogous approach to deal with this issue, but cast it internal to the 
modeling framework to appropriately propagate the uncertainty introduced by additional degrees of 
freedom in the model.  The residual calibration errors are treated as an explicit contribution to 
the model spectrum, enabling us to explore the distribution of possible calibrations (see 
Section~\ref{subsec:MCMC}) and then eventually marginalize out the associated nuisance parameters.  
In essence, the inferences for the relevant stellar parameters will properly account for the 
underlying uncertainty in the calibration process.  This is achieved in practice by distorting the 
model spectrum with a (low-order) Chebyshev polynomial \citep[e.g.,][]{eisenstein06,koleva09}, 
\begin{equation}
\vM(\vt, \vT_{\rm cheb}) = \vM(\vt) \times \sum_n c_n \, T_n,
\label{eqn:chebyshev}
\end{equation}
where $T_n$ are the standard Chebyshev functions of order $n$ and the coefficients are treated as a
set of nuisance (hyper-)parameters, $\vT_{\rm cheb} = \{c_n\}^{\rm order}$, for each available 
spectral order in the dataset.  With judicious priors on $\vT_{\rm cheb}$, we can ensure that the 
unintended treatment of real spectral features (e.g., broad, deep molecular bands) as calibration 
artifacts is negligible (see Section~\ref{sec:examples} for examples).  The lowest-order (scaling) 
coefficient, $c_0$, is by its nature degenerate with the solid angle parameter, $\Omega$.  
Therefore, we enforce an additional constraint by requiring that the mean of the polynomial is 
unity.  For data with a single spectral order, this means setting $c_0 = 1$.  If the goal is to 
model multiple spectral orders, we assign $c_0 = 1$ in an arbitrary order as an anchor, but permit 
the $c_0$ values in other orders to be different (as necessary).  It is worth noting that this 
formalism can, in principle, be extended to develop models for completely uncalibrated spectra, 
rather than the residual calibration mismatches as described here (with a suitable relaxation of 
the priors on $\vT_{\rm cheb}$).  Figure \ref{fig:chebyshev} offers a practical demonstration of how 
these nuisance parameters are applied.

\begin{figure}[!htb]
\begin{center}
  \includegraphics{figs/chebyshev.pdf}
  \caption{The spectrum at the overlap of two echelle orders (23 \& 24). \textbf{Panel 1}: The flux-calibrated dataset shows a slight discrepancy of $\lesssim 3$\% between orders. \textbf{Panel 2}: To account for this residual error in the flux calibration, we multiply the model spectrum by a Chebyshev polynomial, whose coefficients are parameters that are part of the model. \textbf{Panels 3 \& 4}: In principle, we could divide the data by these polynomials to recover what the true flux-calibrated data should be (panel 3), but instead we multiply the model by the polynomials to recover the original discrepancy between orders (panel 4). This has the advantage of preserving the original dataset as a fixed quantity and makes the model (Equation~\ref{eqn:chebyshev}) linear in the Chebyshev polynomial coefficients.}
\label{fig:chebyshev}
\end{center}
\end{figure}

@@

Assuming that these local residuals are produced primarily by pathological differences in the 
spectral line strength (rather than shape or center), a simple Gaussian is a reasonable residual 
model.  In that case, the $k^{\rm th}$ such local residual can be described as
\begin{equation}
\vR_{\lambda}(a_k, \mu_k, \sigma_k) = \frac{a_k}{\sqrt{2\pi}{\sigma_k}} \exp \left[ - \frac{r^2(\lambda,\mu_k)}{2\sigma_k^2} \right],
\end{equation}
with an amplitude $a_k$, a mean wavelength $\mu_k$, and a width $\sigma_k$.  Following 
Eq.~\ref{eqn:expectation}, the kernel that describes the covariance of any two pixels related to 
the $k^{\rm th}$ residual region is
\begin{multline}
  \mathcal{K}_k(\lambda_i, \lambda_j | a_k, \mu_k, \sigma_k) = \\
  \frac{1}{2\pi} \left(\frac{a_k}{\sigma_k}\right)^2 \exp \left [ - \, \frac{r^2(\lambda_i, \mu_k) + r^2(\lambda_j, \mu_k)}{2 \sigma_k^2}\right ].
\label{eqn:klocal}
\end{multline}
We again taper this kernel with a Hann window (Eq.~\ref{eqn:Hann}) to ensure computational 
efficiency with a sparse covariance matrix; in this case, the truncation distance $r_0$ can be set 
to some multiple of the Gaussian width (we set $r_0 = 4\sigma_k$).  
Figure \ref{fig:region} demonstrates how this non-stationary Gaussian 
kernel generates a localized region of enhanced variance that successfully mimics the kind of 
residuals produced by an inaccurate spectral line model.  In effect, these kernels down-weight the 
influence of such strong residuals in the likelihood calculation, mitigating any potential bias 
they might induce on inferences of the interesting parameters ($\vt$).  In essence, this is a 
robust, flexible, and unbiased method for (correlated) outlier rejection that preserves the 
integrity of the probabilistic framework being developed (as opposed to manual or threshold-based 
clipping/masking).

@@

We can now compute the covariance matrix employed in the likelihood calculation 
(Eq.~\ref{eqn:lnprob}) as the linear combination of these kernels and the trivial pixel-by-pixel 
noise matrix, 
\begin{multline}
\vC_{ij}(\vT_{\rm cov})  = b \, \delta_{ij}\sigma_i + \\
         w(r_{ij}|r_0 = 4\ell) \,\, \mathcal{K}_g(\lambda_i, \lambda_j|a_g, \ell) + \\ 
         \sum_k w(r_{ij}|r_0 = 4\sigma_k) \,\, \mathcal{S}_k(\lambda_i,\lambda_j | h_k) \,\, \mathcal{K}_k(\lambda_i, \lambda_j | a_k, \mu_k, \sigma_k),
\end{multline}
where the covariance hyperparameters $\vT_{\rm cov} = [a_g, \ell, \{a_k, \mu_k, \sigma_k, 
h_k\}^{N_{\rm reg}}]$, $N_{\rm reg}$ is the number of local residual regions
(see below for details on how this is determined), and $b$ is assumed to be a
fixed parameter that scales up the pixel noise values to account for read
noise, noise added during the 2D to 1D spectral extraction procedure, and
interpolation uncertainties (see Section \ref{subsec:synthetic}; reasonable
values are $b \approx 1.02$--1.10 for well-calibrated optical spectra; see Section
\ref{sec:examples} for examples).

@@

@@

 We explicitly fit for the
hyperparameters of the covariance kernels at the same time we fit for the
stellar parameters. While including these extra parameters does increase the
dimensionality and complexity of our model, they do provide several advantages.
\emph{A priori} we do not know which regions of the spectrum are improperly
modeled. The covariance hyperparameters provide a framework to identify these
regions iteratively and in a self-consistent manner.

@@

We can (should?) add an additional level to the hierarchy of
  hyperparameters and add parameters to describe the \emph{population
  characteristics} of poorly modeled spectral lines (mostly the typical width,
  amplitude of these lines).  This will tell us about the frequency and
distribution of spectral modelling errors.

@@

\begin{deluxetable}{llrlr@{$\pm$}lr@{$\pm$}lr@{$\pm$}ll} 
\tablecaption{\label{table:tests} Tests}
\tablehead{
  \colhead{Object} & \colhead{Orders} & \colhead{$\lambda$ range [\AA]} & \colhead{library} &
  \multicolumn{2}{c}{$T_{\rm eff} \pm \sigma$} & \multicolumn{2}{c}{$\log g \pm \sigma$} &
  \multicolumn{2}{c}{$\Z \pm \sigma$} & \colhead{comments}
}
\startdata
WASP-14 & 22-24 & 5060-5315 & Kurucz & 6310 & 52 & 3.91  & 0.08 & -0.38  & 0.03 & Poisson only\\
WASP-14 & 22-24 & 5060-5315& Kurucz & 6314 & 50 & 3.92 & 0.08 & -0.39 & 0.03 & Matern only\\
WASP-14 & 22-24 & 5060-5315& Kurucz & 6512 & 32 & \multicolumn{2}{c}{\nodata} & -0.27 & 0.02 & Matern + fixed $\log g = 4.29$\\
WASP-14 & 23 & & PHOENIX & 6021 & 16 & 3.83 & 0.03 & -0.50 & 0.01 & Poisson only \\
WASP-14 & 23 & & PHOENIX & 5965 & 70 & 3.78 & 0.09 & -0.68 & 0.07 & Matern only, no regions \\
WASP-14 & 23 & & PHOENIX & & & & & & & Matern and regions \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & 6117 & 30 & 3.73 & 0.06 & -0.52 & 0.02 & Poisson only \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & 5865 & 43 & 3.2 & 0.08 & -0.85 & 0.03 & Matern only. Because of 24. \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & & & \multicolumn{2}{c}{\nodata} & & & Matern + fixed $\log g = 4.29$ \\
WASP-14 & 22-24 & 5060-5315 & PHOENIX & & & & & & & Regions included \\
\enddata
\end{deluxetable}
